---
format: html
---

## Applied Mathematics in Biology: Single Cell Omics

Lecture Notes

Simon Anders, Univ. Heidelberg, 2023-06

## Pseudotime

### Today's example data

This time, we use a slightly different data set, from another experiment but with a similar setting,
namely again GLAST+ cells from the murine sSVZ, this time comparing wildtype with a double-knockout
of both interferon receptors (alpha/beta and gamma).

The data was the basis of this publication: Carvajal Ibañez, Skabkin, Hooli, et al.: *Interferon regulates neural stem cell function at allages by orchestrating mTOR and cell cycle*. EMBO Molecular Medicine 15:e16434 (2023).

### Loading and preprocessing

Here, we perform preprocessing as before.

```{r}
suppressPackageStartupMessages({
  library( tidyverse )
  library( Matrix ) })
```

The data is provided, as before, in MatrixMarket format:

```{r}
mtx_mm <- read.delim( gzfile( "/mnt/raid/ifnagrko/ifnagrko_raw_counts.mtx.gz" ), 
   comment.char="%", sep=" ", header=FALSE )
head( mtx_mm )
```

Again, we have a seperate file with gene annotations. We remove the first column, as
it contains only line numbers.

```{r}
genes <- read_csv( "/mnt/raid/ifnagrko/ifnagrko_var.csv", show_col_types=FALSE )[,-1] 
head( genes )
```
For the cell annotation, our file is richer than usual as it contains not
only the barcodes (as usual) but also all the results of Damian et al.'s 
previous analysis:

```{r}
cells <- read_csv( "/mnt/raid/ifnagrko/ifnagrko_obs.csv" )
head(cells)
```
We assemble the count matrix:

```{r}
counts <- 
  sparseMatrix(
    dims = c( mtx_mm$V2[1], mtx_mm$V1[1] ),
    i = mtx_mm$V2[-1],
    j = mtx_mm$V1[-1],
    x = mtx_mm$V3[-1],
    dimnames = list( genes$gene_name, cells$barcode ) )

counts[ 1:5, 1:5 ]
```  

Free up some memory
```{r}
rm( mtx_mm )
```

Log-normalization:
```{r}
totals <- colSums(counts)
fracs <- t( t(counts) / totals )
lfracs <- log1p( fracs * 1e4 )
```

Selection of highly variable genes:

```{r}
gene_frac_means <- rowMeans( fracs )
gene_frac_vars <- sparseMatrixStats::rowVars( fracs )
sort( gene_frac_vars / gene_frac_means, decreasing=TRUE ) %>% 
   head( 1000 ) %>% names() -> hvg

plot( gene_frac_means, gene_frac_vars / gene_frac_means, pch=".", log="xy",
      col = ifelse( names(gene_frac_means) %in% hvg, "red", "black" ))
abline( h = mean( 1/totals), col="darkblue" )
```

Perform PCA:

```{r}
pca <- irlba::prcomp_irlba( t(lfracs[ hvg, ]), n=30, scale.=TRUE )
rownames(pca$x) <- colnames(lfracs)   # cells
rownames(pca$rotation) <- hvg         # genes
```

Plot first two PCs

```{r}
plot( pca$x[,1], pca$x[,2], pch=".", asp=1 )
```
Get a UMAP. This time, we let `umap` do the neighborhood search and extract the result
for further use
```{r}
ump <- uwot::umap( pca$x, min_dist=.1, ret_nn=TRUE )
nn <- ump$nn$euclidean
ump <- ump$embedding
```

Here is the UMAP:
```{r}
plot( ump, pch=".", asp=1, col = adjustcolor( "black", alpha=.2 ) )
```

It's useful to use sleepwalk, too. (Not run.)
```{r eval=FALSE}
sleepwalk::sleepwalk( ump, pca$x )
```

For Leiden clustering, we get the NN graph from the NN data returned by umap,
then run `igraph::leiden`:

```{r}
edge_table <-
  map_dfr( 2:ncol(nn$idx), function(i)
    tibble( 
      from = 1:nrow(nn$idx), 
      to = nn$idx[,i] ) )
nn_graph <- igraph::graph_from_edgelist( as.matrix(edge_table), directed=FALSE )
set.seed( 13245768 )
leiden <- igraph::cluster_leiden( nn_graph, objective_function="modularity" )
```

Here's the clusters:

```{r}
as_tibble( ump ) %>%
  add_column( cluster = factor( leiden$membership) ) -> ump_with_cluster

ump_with_cluster %>%
  group_by( cluster ) %>%
  summarise_all( mean ) -> cluster_centers_ump

ggplot( ump_with_cluster, aes( x=V1, y=V2 ) ) +
  geom_point( aes( col=cluster ), size=.1 ) +
  geom_text( aes( label=cluster ), data=cluster_centers_ump ) +
  coord_equal()
```
Damian et al. (the paper authors) have performed clustering, too, and annotated 
the clusters with cell types. We plot their annotation into our UMAP, for comparison:

```{r}
as_tibble( ump ) %>%
bind_cols( cells ) %>%
group_by( celltype ) %>%
summarise( V1=mean(V1), V2=mean(V2) ) -> tmp

as_tibble( ump ) %>%
bind_cols( cells ) %>%
ggplot( aes( x=V1, y=V2 ) ) +
  geom_point( aes( col=celltype ), size=.1 ) +
  geom_text( aes( label=celltype ), data = tmp, angle=-30 ) +
  coord_equal() +
  guides( colour = guide_legend( override.aes = list(size=3) ) )
```

Of interest for us is the "lineage", i.e., teh progression from quiescent neural stem cells (qNSCs)
via active stem cells (aNSCs) and transit amplifying progenitors (TAPs) towards neuroblasts (NBs).
We can only see early NBs and no late NBs are mature neurons as the cells have been gated for
GLAST protein expression. 

Here are these "lineage" cells all highlighted in one color:

```{r}
as_tibble( ump ) %>%
bind_cols( cells ) %>%
ggplot( aes( x=V1, y=V2 ) ) +
geom_point( aes( col=celltype1 ), size=.1 ) +
  coord_equal() +
  guides( colour = guide_legend( override.aes = list(size=3) ) )
```

### Pseudotime

Within the lineage, there is a gradual, continuous transition, and it using
Leiden clustering, which imposes sharp boundaries, does not seem the right 
way. Hence, we want instead to assign to each lineage cell a *pseudotime* -- 
a value indicating how far along the cell is in the progression.

Here is the pseudotime used in the paper:

```{r}
as_tibble( ump ) %>%
bind_cols( cells ) %>%
ggplot( aes( x=V1, y=V2 ) ) +
geom_point( aes( col=dpt_pseudotime ), size=.1 ) +
  coord_equal() +
  scale_color_viridis_c()
```
They used a method known as "diffusion pseudotime" (DPT), implemented in the "destiny"
package, to get there. We will first try another method, known as "principal curve".

First, however, just for comparison, again *their* pseudotime, but now with *their*
UMAP:

```{r}
cells %>%
ggplot( aes( x=UMAP1, y=UMAP2 ) ) +
geom_point( aes( col=dpt_pseudotime ), size=.1 ) +
  coord_equal() +
  scale_color_viridis_c()
```

### Principal curves

The "princurve" package allow to fit a "principal curve" through a point cloud.
This is a curve chosen such that the variance of the points images of a normal
projection onto the curve is maximal (with some constraints on the curve to get
some "stiffness").

We try this out. As the function is slow, we use only a subset (sample) of the lineage cells.

```{r}
set.seed( 13245768 )
sampled_lng_cells <- sample( cells$barcode[ cells$celltype1=="lineage" ], 2000 )
prcs <- princurve::principal_curve( pca$x[ sampled_lng_cells, ], df=10, trace=TRUE )
```
Here's the plot:

```{r}
as_tibble( ump, rownames="cell" ) %>%
left_join( enframe( prcs$lambda, "cell", "lambda" ) ) %>%
arrange( !is.na(lambda) ) %>%
ggplot( aes( x=V1, y=V2 ) ) +
  geom_point( aes( col=lambda ), size=.1 ) +
  coord_equal() + scale_color_viridis_c()
```
How do the two pseudotimes compare?

```{r}
cells %>%
left_join( enframe( prcs$lambda, "barcode", "lambda" ) ) %>%
ggplot() +
  geom_point( aes( x=dpt_pseudotime, y=lambda ), size=.4 )
```

There seems to be noticeable disagreement.

I am, to be honest, not sure which result to believe more, and therefore will 
perform a few more analyses. For now, let's use the pseudotime from the paper.

### Expression along pseudotime

How does the expression of a gene change along the pseudotime?

Let's look at Slc1a3, the gene coding for GLAST:

First, a normal coloured UMAP feature plot:

```{r}
as_tibble( ump ) %>%
add_column( expr = lfracs["Slc2a1",] ) %>%
ggplot( aes( x=V1, y=V2 ) ) +
  geom_point( aes( col=expr ), size=.1, alpha=.3 ) +
  coord_equal() + scale_color_viridis_c( direction=-1 )
```
Maybe, binarizing it makes it easier to see:

```{r}
as_tibble( ump ) %>%
add_column( expr = lfracs["Slc2a1",]>0 ) %>%
ggplot( aes( x=V1, y=V2 ) ) +
  geom_point( aes( col=expr ), size=.1, alpha=.3 ) +
  coord_equal() + scale_color_manual( values=c("gray80","darkblue") )
``` 

Still, hard to see something. Typical "salt-and-pepper plot".

Plotting against pseudotime might make things clearer:

```{r}
plot( cells$dpt_pseudotime, lfracs["Slc2a1",])
```

As there are so many zero values, this is still hard to interpret.

Let's sort the cells by pseudotime, form "bins" of 50 cells each that
are next to each other in pseudotime, and average:

```{r}
tibble( 
  pt = cells$dpt_pseudotime,
  expr = lfracs["Slc2a1",] ) %>%
filter( !is.na(pt) ) %>%
arrange( pt ) %>%
mutate( pt_bin = floor( row_number() / 50 ) ) %>%
group_by( pt_bin ) %>%
summarise_at( c( "pt", "expr" ), mean ) %>%
ggplot() +
  geom_point( aes( x=pt, y=expr ) )
```

This plot has two problems:

- It's still not a smooth curve
- The distance of the zero line from the rest depends on the scaling factor we used in the 
  log-normalisation.
  
To see the latter, change teh scaling factor from 1e4 to 1e5:  

```{r}
tibble( 
  pt = cells$dpt_pseudotime,
  expr = log1p( fracs["Slc2a1",] * 1e5 ) ) %>%
filter( !is.na(pt) ) %>%
arrange( pt ) %>%
mutate( pt_bin = floor( row_number() / 50 ) ) %>%
group_by( pt_bin ) %>%
summarise_at( c( "pt", "expr" ), mean ) %>%
ggplot() +
  geom_point( aes( x=pt, y=expr ) )
```

Let's use the "locfit" package to get a decent smooth curve:

```{r}
library( locfit )

tibble( 
  t = cells$dpt_pseudotime,
  k = counts["Slc2a1",],
  s = totals ) %>%
filter( !is.na(t) ) -> ptfit_data

fit <- locfit( k ~ lp(t,h=.2), base=log(s), ptfit_data, family="poisson", ev=lfgrid(300) )
plot( fit, log="" )
```

To understand what is going on here, we need to learn about *kernel smoothing* and
*local regression*. For this, we start a new notebook.

### Homework 1

Find 20 genes that differ strongly between qNSCs and NBs. Calculate smoothed fits 
of their expression along pseudotime (like the one just made above). Make a heatmap
where the rows are genes, the columns are pseudotime bins, and the colour is the fitted
value for the gene's expression at the respective pseudotime points. 

Compare a heatmap made with locfit with one made by simple pseudotime binning and 
averaging.

### Homework 2

Above, we have averaged over cells with similar pseudotime. So, we have reduced
our analysis to a 1D look. Let's try something different but no more difficult:

For each cell, get three estimates of its expression of Slc1a3:

- First, the usual lfrac value, i.e., $\log( k/s \cdot 10^4 + 1)$.

- Second: Average these lfrac values over the cell's 20 (or 50) nearest neighbors.
  Include the cell itself among its neighbours. Consider this average-over-neighbors
  as an estimate of the cell's expression. 
  
- Third: Instead of averaging the log-fractions, add up the counts for the gene of interest
  over all the neighbours (including the cell itself) and divide it by the sum of all the total,
  i.e., calculate $\sum_j k_j / \sum_j s_j$, where both sums run over the cell and its 20
  nearest neighbors.
  
Plot UMAPs and color the points by these estimates. Experiment with different neighborhood
sizes. Which seems to be most useful?

Also: Calculate for each cell the "deviance" between of its neighborhood-smoothed estimate $\hat q$
and its observed value $k$:

$$ D = -2 \left( \log f_\text{Pois}( k; s\hat q ) -  \log f_\text{Pois}( k; k ) \right), $$
where $f_\text{Pois}(k,\lambda)$ is the Poisson probability mass function, i.e., the probability
to observe $k$ when drawing from a Poisson distribution with rate parameter $\lambda$. (In R, use `dpois( k, lambda, log=TRUE )` to get $\log f_\text{Pois}(k,\lambda)$.) 

Derive from the deviance the "deviance residual" $z=\sqrt{D}\cdot\operatorname{sgn}(k-s\hat q)$ and use that to colour
your UMAPs. What might this tell you?
